\section{Conclusion}
\label{sec:conclusion}

We have presented SuperLocalMemory V3.3, the first local-first AI agent memory system implementing the complete cognitive memory taxonomy. Five contributions advance the state of the art:

\begin{enumerate}[topsep=2pt, itemsep=1pt]
  \item \textbf{FRQAD} achieves 100\% precision at distinguishing full-fidelity from quantized embeddings, with zero prior art.
  \item \textbf{Ebbinghaus Adaptive Forgetting} provides 6.7$\times$ discriminative power with provable convergence.
  \item \textbf{7-Channel Cognitive Retrieval} achieves 70.4\% on LoCoMo in zero-LLM mode, with +23.8pp on multi-hop.
  \item \textbf{Memory Parameterization} implements Long-Term Implicit memory at zero LLM cost.
  \item \textbf{Zero-Friction Pipeline} automates the complete memory lifecycle via a single install.
\end{enumerate}

The mathematics are sound---the Fokker-Planck system converges, FRQAD is a valid metric on the quantized Gaussian manifold, and the forgetting-quantization coupling is self-consistent with the Fisher-Rao geometry. The engineering is production-grade---3,000+ tests, deployed on npm and PyPI with over 5,000 monthly downloads.

For the first time, a developer can install a single package and receive automatic, privacy-preserving cognitive memory across every AI coding session. No cloud. No API keys. No manual commands.

\emph{Every other AI forgets. Yours won't.}

\vspace{12pt}
\noindent\textbf{Availability.} We release all code under the Elastic License~2.0 for reproducibility at \url{https://github.com/qualixar/superlocalmemory}. The system is available via npm (\texttt{superlocalmemory}) and PyPI (\texttt{superlocalmemory}). Documentation: \url{https://superlocalmemory.com}. Zenodo DOI: \url{https://doi.org/10.5281/zenodo.19435120}.

% ============================================================================
% ACKNOWLEDGMENTS
% ============================================================================
