% CVPR 2026 Paper Template; see https://github.com/cvpr-org/author-kit

\documentclass[10pt,twocolumn,letterpaper]{article}

%%%%%%%%% PAPER TYPE  - PLEASE UPDATE FOR FINAL VERSION
\usepackage{cvpr}              % To produce the CAMERA-READY version
% \usepackage[review]{cvpr}      % To produce the REVIEW version
% \usepackage[pagenumbers]{cvpr} % To force page numbers, e.g. for an arXiv version

% Import additional packages in the preamble file, before hyperref
\input{preamble}

% It is strongly recommended to use hyperref, especially for the review version.
% hyperref with option pagebackref eases the reviewers' job.
% Please disable hyperref *only* if you encounter grave issues, 
% e.g. with the file validation for the camera-ready version.
%
% If you comment hyperref and then uncomment it, you should delete *.aux before re-running LaTeX.
% (Or just hit 'q' on the first LaTeX run, let it finish, and you should be clear).
\definecolor{cvprblue}{rgb}{0.21,0.49,0.74}
\usepackage[pagebackref,breaklinks,colorlinks,allcolors=cvprblue]{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{paralist}
\usepackage{wrapfig}
% \usepackage[shortlabels,inline]{enumitem}

%%%%%%%%% PAPER ID  - PLEASE UPDATE
\def\paperID{18659} % *** Enter the Paper ID here
\def\confName{CVPR}
\def\confYear{2026}

%%%%%%%%% TITLE - PLEASE UPDATE
% \title{You Can Tokenize Longer fMRI Brain Sequences Using Off-the-shelf Natural Image Autoencoders?}
% \title{Can Natural Image Autoencoders Tokenize fMRI Tokenization for Long-Range Dynamics Modeling?}
\title{Can Natural Image Autoencoders Compactly Tokenize fMRI Volumes \\for Long-Range Dynamics Modeling?}
% \title{2D Natural Image Autoencoder can Model Long-range fMRI Dynamics}

% Can Natural Image Autoencoders Tokenize 3D fMRI for Long-Range Dynamics Modeling?
% Can Natural Image Autoencoders Compactly Tokenize fMRI for Long-Range Dynamics Modeling?
% Compact fMRI Tokenization with Natural Image Autoencoders for Long-Range Dynamics Modeling

% Compact tokenization of long-range fMRI signals via natural image autoencoders

% By 정우
% Can Pretrained 2D Natural Image Autoencoders Tokenize fMRI Volumes into Just 27 Tokens?

% Are Off-the-Shelf 2D Image Autoencoders Surprisingly Superior Tokenizers for fMRI Dynamics?
%%%%%

% Achieving Efficient Spatio-temporal Trade-off for  Long-Range fMRI Dynamics Modeling with Natural Image Autoencoder


%%%%%%%%% AUTHORS - PLEASE UPDATE
\author{Peter Yongho Kim$^{1}$\thanks{Equal contribution.} , Juhyeon Park$^{2}$\footnotemark[1] , Jungwoo Park$^{1}$, \\Jubin Choi$^{2}$, Jungwoo Seo$^{3}$, Jiook Cha$^{2,4}$, Taesup Moon$^{1,2,5}$\thanks{Corresponding author.} \\
$^{1}$ECE, Seoul National University,
$^{2}$IPAI, Seoul National University, \\
$^{3}$BCS, Seoul National University,
$^{4}$Psychology, Seoul National University, \\
$^{5}$ASRI / INMC / AIIS, Seoul National University\\
\texttt{\{peterkim98, parkjh9229, tsmoon\}@snu.ac.kr} \\
}
% \author{Peter Yongho Kim\thanks{Equal contribution.} , Juhyeon Park\footnotemark[1] , Jungwoo Park, Jubin Choi, Jungwoo Seo, Jiook Cha, Taesup Moon\thanks{Corresponding author.}\\
% Seoul National University\\
% % Institution1 address\\
% % {\tt\small firstauthor@i1.org}
% % For a paper whose authors are all at the same institution,
% % omit the following lines up until the closing ``}''.
% % Additional authors and addresses can be added with ``\and'',
% % just like the second author.
% % To save space, use either the email address or home page, not both
% % \and
% % Second Author\\
% % Institution2\\
% % First line of institution2 address\\
% % {\tt\small secondauthor@i2.org}
% }
\newcommand{\ours}{TABLeT}
\newcommand{\meanstd}[2]{$#1_{\pm #2}$}
\newcommand{\RNum}[1]{\uppercase\expandafter{\romannumeral #1\relax}}

% \usepackage{nameref,zref-xr}
% \zxrsetup{toltxlabel}
% \zexternaldocument*{supple_for_main_sep}

\begin{document}
\maketitle


\begin{abstract}
Modeling long-range spatiotemporal dynamics in functional Magnetic Resonance Imaging (fMRI) remains a key challenge due to the high dimensionality of the four-dimensional signals. 
Prior voxel-based models, although demonstrating excellent performance and interpretation capabilities, are constrained by prohibitive memory demands and thus can only capture limited temporal windows. 
To address this, we propose TABLeT (Two-dimensionally Autoencoded Brain Latent Transformer), a novel approach that tokenizes fMRI volumes using a pre-trained 2D natural image autoencoder. Each 3D fMRI volume is compressed into a compact set of continuous tokens, enabling long-sequence modeling with a simple Transformer encoder with limited VRAM. Across large-scale benchmarks including the UK-Biobank (UKB), Human Connectome Project (HCP), and ADHD-200 datasets, TABLeT outperforms existing models in multiple tasks, while demonstrating substantial gains in computational and memory efficiency over the state-of-the-art voxel-based method given the same input. 
Furthermore, we develop a self-supervised masked token modeling approach to pre-train {\ours}, which improves the model's performance for various downstream tasks.
Our findings suggest a promising approach for scalable and interpretable spatiotemporal modeling of brain activity.
Our code is available at \url{https://github.com/beotborry/TABLeT}.
\end{abstract}


\input{contents/introduction}
\input{contents/related_works}
\input{contents/method}
\input{contents/experimental_results}
\input{contents/conclusion}
\section*{Acknowledgment}
This work was supported in part by National Research Foundation of Korea (NRF) grant
[No. RS-2025-02263628, No. RS-202300265406], the Institute of Information \& communications Technology Planning
\& Evaluation (IITP) grants [RS-2021-II212068, RS-2022-II220113, RS-2022-II220959, RS-2021-II211343], the BK21 FOUR Education and
Research Program for Future ICT Pioneers (Seoul
National University), funded by the Korean government (MSIT), the Ministry of Education [RS-2024-00435727], the National Supercomputing Center [KSC-2025-CRE-0340], the U.S. Department of Energy’s ASCR Leadership Computing Challenge [m4750-2024], and Hyundai Motor Chung Mong-Koo Foundation.


{
    \small
    \bibliographystyle{ieeenat_fullname}
    \bibliography{main}
}

% WARNING: do not forget to delete the supplementary pages from your submission 
% \input{sec/X_suppl}
\appendix
\input{contents/supplementary}

\end{document}
